# R Basics

Before we get started with this analysis we are reviewing some basics.


## Packages

Use `install.packages` to install the **dslabs** package.

Tryout the following functions: `sessionInfo`, `installed.packages`

## Prebuilt functions

Much of what we do in R is called prebuilt functions. Today we are using: `ls`, `rm`, `library`, `search`, `factor`, `list`, `exists`, `str`, `typeof`, `class` and maybe more.

You can see the code for a function by typing it without the parenthesis:

Try this:
```{r}
#| eval: false

ls
```

## Help system

In R you can use `?` or `help` to learn more about functions.

You can learn about function using

```
help("ls")
```

or 

```
?ls
```

## The workspace

Define a variable.

Use `ls` to see if it's there. Also take a look at the **Environment** tab in RStudio.

Use `rm` to remove the variable you defined.



## Variable name convention

A nice convention to follow is to use meaningful words that describe what is stored, use only lower case, and use underscores as a substitute for spaces. 

For more we recommend [this guide](http://adv-r.had.co.nz/Style.html).

## Data types

The main data types in R are:

* One dimensional vectors: numeric, integer, logical, complex, characters.

* Factors

* Lists: this includes data frames

* Arrays: Matrices are the most widely used

* Date and time

* tibble

* S4 objects

Many errors in R come from confusing data types. Let's learn what these data types are and useful tools to help us.

`str` stands for structure, gives us information about an object.

`typeof` gives you the basic data type of the object. It reveals the lower-level, more fundamental type of an object in R's memory.

`class` This function returns the class attribute of an object. The class of an object is essentially `type_of` at a higher, often user-facing level.

```{r}
library(dslabs)
```


```{r}
typeof(murders)
```


```{r}
class(murders)
```


```{r}
str(murders)
```


## Data frames

Date frames are the most common class used in data analysis. It is like a spreadsheet. Rows represents observations and columns variables. Each variable can be a different data type.

You can add columns like this:

```{r}
murders$pop_rank <- rank(murders$population)
head(murders)
```

You can access columns with the `$` 

```{r}
murders$population
```

but also `[]`
```{r}
murders[1:5,]
```

```{r}
murders[1:5, 1:2]
```

```{r}
murders[1:5, c("state", "abb")]
```

## with

The function `with` let's us use the column names as objects:

```{r}
with(murders, length(state))
```


## Vectors

The columns of data frames are one dimensional (atomic) vectors.


Here is an example:
```{r}
length(murders$population)
```


How to create vectors:

```{r}
x <- c("b", "s", "t", " ", "2", "6", "0")
```

Sequences are particularly useful:

```{r}
seq(1, 10)
```

```{r}
seq(1, 9, 2)
```

```{r}
1:10
```

```{r}
seq_along(x)
```

## Factors

One key data type distinction is factors versus characters:

```{r}
typeof(murders$state)
typeof(murders$region)
```

Factors store levels and then the label of each level.
They are very useful for categorical data.

```{r}
x <- murders$region
levels(x)
```


### Categories based on strata

The function `cut` is useful for converting numbers into categories

```{r}
with(murders, cut(population, 
                  c(0, 10^6, 10^7, Inf)))
```

```{r}
murders$size <- cut(murders$population, c(0, 10^6, 10^7, Inf), 
            labels = c("small", "medium", "large"))
murders[1:6,c("state", "size")]
```


### changing levels

You can change the levels (this will come in handy when we learn linear models)

Order levels alphabetically:
```{r}
factor(x, levels = sort(levels(murders$region)))
```

Make west the first level:
```{r}
x <- relevel(x, ref = "West")
```

Order levels by population size:
```{r}
x <- reorder(murders$region, murders$population, sum)
```

Factors are more efficient:

```{r}
x <- sample(murders$state[c(5,33,44)], 10^7, replace = TRUE)
y <- factor(x)
object.size(x)
object.size(y)
```

```{r}
system.time({x <- tolower(x)})
```

Exercise: How can we make this go much faster?

```{r}
system.time({levels(y) <- tolower(levels(y))})
```

Factors can be confusing:

```{r}
x <- factor(c("3","2","1"), levels = c("3","2","1"))
as.numeric(x)
```


```{r}
x[1]
levels(x[1])

table(x[1])

z <- x[1]
z <- droplevels(z)
```

```{r}
x[1] <- "4"
x
```


## NAs

NA stands for not available. We will see many NAs if we analyze data generally.

```{r}
x <- as.numeric("a")
```

```{r}
is.na(x)
```

```{r}
is.na("a")
```

```{r}
1 + 2 + NA
```

When used with logicals behaves like FALSE
```{r}
TRUE & NA
TRUE | NA
```

But is is not FALSE. Try this:

```{r}
#| eval: false
if (NA) print(1) else print(0)
```

A related constant is `NaN` which stands for not a number. It is a numeric that is not a number.

```{r}
class(0/0)
sqrt(-1)
log(-1)
0/0
```


## coercing

When you do something nonsensical with data types, R tries to figure out what you mean. This can cause confusion and unnoticed errors. So it's important to understand how and when it happens. 
Here are some examples:

```{r}
typeof(1L)
typeof(1)
typeof(1 + 1L)
```

```{r}
c("a", 1, 2)
```

```{r}
TRUE + FALSE
```

```{r}
factor("a") == "a"
identical(factor("a"), "a")
```

You want to avoid automatic coercion and instead explicitly do it. Most coercion functions start with `as.`

```{r}
x <- factor(c("a","b","b","c"))
as.character(x)
as.numeric(x)
```

```{r}
x <- c("12323", "12,323")
as.numeric(x)
readr::parse_guess(x)
```

## lists

Data frames are a type of list. List permit components of different types and, unlike data frames, length

```{r}
x <- list(name = "John", id = 112, grades = c(95, 87, 92))
```

You can access components in different ways:

```{r}
x$name
x[[1]]
x[["name"]]
```

## matrics

Matrices are another widely used data type. They are similar to data frames except all entries need to be of the same type.

We will learn more about matrices in the High Dimensional data Analysis part of the class.

## functions

You can define your own function. The form is like this:

```{r}
#| eval: false
f <- function(x, y, z = 0){
  ### do calculations with x, y, z to compute object
  ## return(object)
}
```

Here is an example of a function that sums $1,2,\dots,n$

```{r}
s <- function(n){
   return(sum(1:n))
}
```

## Lexical scope

```{r}
f <- function(x){
  cat("y is", y,"\n")
  y <- x
  cat("y is", y,"\n")
  return(y)
}
y <- 2
f(3)
y <- f(3)
y
```


## Namespaces 


Look at this function.


```{r}
#| eval: false

filter
library(dplyr)
filter
```


Note this is just the Global Environment.

Use `search` to see other environments.

Note all the functions in `stats`

You can explicitly say which you want:

```{r}
#| eval: false
stats::filter
dplyr::filter
```

Try to understand this example:

```{r}
exists("murders")
library(dslabs)
exists("murders")
murders <- murders
murders2 <- murders
rm(murders)
exists("murders")
detach("package:dslabs")
exists("murders")
exists("murders2")
```

## object oriented programming

R uses object oriented programming. It uses to approaches referred to as S3 and S4. The original S3 is more common.

What does this mean?

```{r}
#| label: co2-plot
class(co2)
plot(co2)
```

```{r}
#| label: wrong-co2-plot
plot(as.numeric(co2))
```

See the difference? The first one actually calls the function

```{r}
#| eval: false
plot.ts
```

Notice all the plot functions that start with `plot.` 

The function plot will call different functions depending on the class of the arguments:

```{r}
plot
```


## Exercises

(@) What is the sum of the first 100 positive integers? The formula for the sum of integers $1$ through $n$ is $n(n+1)/2$. Define $n=100$ and then use R to compute the sum of $1$ through $100$ using the formula. What is the sum?

```{r}
n <- 100
n*(n + 1) / 2
```

(@) Now use the same formula to compute the sum of the integers from 1 through 1,000.

```{r}
n <- 1000
n*(n + 1) / 2
```


(@) Now use the functions `seq` and `sum` to compute the sum with R for any `n`, rather than a formula.

```{r}
n <- 100
x <- seq(1, 100)
sum(x)
```

(@) In math and programming, we say that we evaluate a function when we replace the argument with a given number. So if we type `sqrt(4)`, we evaluate the `sqrt` function. In R, you can evaluate a function inside another function. The evaluations happen from the inside out. Use one line of code to compute the log, in base 10, of the square root of 100.

```{r}
log(sqrt(100), base = 10)
log10(sqrt(100))
```

(@) Make sure the US murders dataset is loaded. Use the function `str` to examine the structure of the `murders` object. What are the column names used by the data frame for these five variables?

```{r}
library(dslabs)
str(murders)
```

