# Distributions 

## Case study: describing student heights

We will study self reported heights from studnets from past classes:

```{r load-heights, warning=FALSE, message=FALSE}
library(tidyverse)
library(dslabs)
head(heights)
```


## Distributions

The most basic statistical summary of a list of objects or numbers is its *distribution*. 


```{r}
prop.table(table(heights$sex))
```

Here is the distribution for the regions in the `murders` dataset:


```{r state-region-distribution, echo=FALSE}
murders |> group_by(region) |>
  summarize(n = n()) |>
  mutate(Proportion = n/sum(n), 
         region = reorder(region, Proportion)) |>
  ggplot(aes(x = region, y = Proportion, fill = region)) + 
  geom_col(show.legend = FALSE) + 
  xlab("")
```


### Histograms

Cumulative distributions function shows everything you need to know the distribution.

```{r ecdf, echo=FALSE}
ds_theme_set()
heights |> filter(sex == "Male") |> ggplot(aes(height)) + 
  stat_ecdf() +
  ylab("Proportion of heights less than or equal to a") + 
  xlab("a")
```

Histograms lose a bit fo information but are easier to read:

```{r height-histogram, echo=FALSE}
heights |> 
  filter(sex == "Male") |> 
  ggplot(aes(height)) + 
  geom_histogram(binwidth = 1, color = "black")
```


### Smoothed density

*Smooth density* plots relay the same information as a histogram but are aesthetically more appealing. Here is what a smooth density plot looks like for our heights data:

```{r example-of-smoothed-density, echo=FALSE}
heights |> 
  filter(sex == "Male") |> 
  ggplot(aes(height)) + 
  geom_density(alpha = .2, fill = "#00BFC4", color = 0)  +
  geom_line(stat = 'density')
```

An advantage is that it is easy to show more than one:

```{r two-densities-one-plot, echo=FALSE}
heights |> 
  ggplot(aes(height, fill = sex)) + 
  geom_density(alpha = 0.2, color = 0) +
  geom_line(stat = 'density')
```

With the right argument, `ggplot` automatically shades the intersecting region with a different color. 

### The normal distribution {#sec-dataviz-normal-distribution}


The normal distribution, also known as the bell curve and as the Gaussian distribution. Here is what the normal distribution looks like:

```{r normal-distribution-density, echo=FALSE}
mu <- 0; s <- 1
norm_dist <- data.frame(x = seq(-4,4, len = 50)*s + mu) |> mutate(density = dnorm(x, mu, s))
norm_dist |> ggplot(aes(x,density)) + geom_line()
```

A useful characteristic of the normal distribution is that it is defined by just two numbers: the average (also called mean) and the standard deviation.

So for the male height data we can define the average of standard deviation like this:

```{r}
index <- heights$sex == "Male"
x <- heights$height[index]
m <- sum(x) / length(x)
s <- sqrt(sum((x - mu)^2)/length(x))
```

The pre-built functions `mean` and `sd` can be used here:
:::

```{r}
m <- mean(x)
s <- sd(x)
c(average = m, sd = s)
```

:::{.callout-note}
The pre-built functions `mean` and `sd` (note that, for reasons explained in statistics textbooks,`sd` divides by `length(x)-1` rather than `length(x)`) can be used here:
:::

Here is a plot of the smooth density and the normal distribution with mean = `r round(m,1)` and SD = `r round(s,1)` plotted as a black line with our student height smooth density in blue:

```{r data-and-normal-densities, echo=FALSE}
norm_dist <- data.frame(x = seq(-4, 4, len = 50)*s + m) |> 
  mutate(density = dnorm(x, m, s))

heights |> filter(sex == "Male") |> ggplot(aes(height)) +
  geom_density(fill = "#0099FF") +
  geom_line(aes(x, density),  data = norm_dist, lwd = 1.5) 
```

## Boxplots

Boxplots provide a five number summary (and shows outliers):

```{r hist-non-normal-data, echo=FALSE, message=FALSE}
murders <- murders |> mutate(rate = total/population*100000)
library(gridExtra)
murders |> ggplot(aes(x = rate)) + geom_histogram(binwidth = 0.5, color = "black") + ggtitle("Histogram")
```

In this case, the histogram above or a smooth density plot would serve as a relatively succinct summary.


```{r first-boxplot, echo=FALSE}
murders |> ggplot(aes("",rate)) + geom_boxplot() +
  coord_cartesian(xlim = c(0, 2)) + xlab("")
```


## Stratification {#sec-dataviz-stratification}

Showing _conditional_ distributions is often very informative

```{r female-male-boxplots, echo=FALSE}
heights |> ggplot(aes(x = sex, y = height, fill = sex)) +
  geom_boxplot()
```


We also see the normal approximation might not be useful for females:

```{r histogram-qqplot-female-heights, echo=FALSE}
heights |> filter(sex == "Female") |>
  ggplot(aes(height)) +
  geom_density(fill = "#F8766D") 
```


Regarding the five smallest values, note that these values are:

```{r}
heights |> filter(sex == "Female") |> 
  top_n(5, desc(height)) |>
  pull(height)
```

Because these are reported heights, a possibility is that the student meant to enter `5'1"`, `5'2"`, `5'3"` or `5'5"`.

## Exercises


(@) Suppose we can't make a plot and want to compare the distributions side by side. We can't just list all the numbers. Instead, we will look at the percentiles. Create a five row table showing `female_percentiles` and `male_percentiles` with the 10th, 30th, 50th, 70th, & 90th percentiles for each sex. Then create a data frame with these two as columns.

```{r}
library(dslabs)
## Here is an R-base solution
qs <- seq(10,90,20)
with(heights,
     data.frame(
       quantile(height[sex == "Male"], qs/100),
       quantile(height[sex == "Female"], qs/100)
)) |> setNames(c("female_percentiles", "male_percentiles"))

## Here is the solution using pivot_wider, which we learn later
library(dplyr)
qs <- seq(10,90,20)
heights |> group_by(sex) |>
  reframe(quantile = paste0(qs, "%"), value = quantile(height, qs/100)) |>
  pivot_wider(names_from = sex) |>
  rename(female_percentiles = Female, male_percentiles = Male)
```

(@) Study the following boxplots showing population sizes by country:

```{r boxplot-exercise, echo=FALSE, message = FALSE}
library(tidyverse)
library(dslabs)
ds_theme_set()
tab <- gapminder |> filter(year == 2010) |> group_by(continent) |> select(continent, population)  
tab |> ggplot(aes(x = continent, y = population/10^6)) + 
  geom_boxplot() + 
  scale_y_continuous(trans = "log10", breaks = c(1,10,100,1000)) + ylab("Population in millions")
```

Which continent has the country with the biggest population size?

(@) What continent has the largest median population size?

(@) What is median population size for Africa to the nearest million?

(@) What proportion of countries in Europe have populations below 14 million?

a.  0.99
b.  0.75
c.  0.50
d.  0.25

(@) When using the log transformation, which continent shown above has the largest interquartile range?

```{r}
## We can see that it is Americas visually, but just in case here it is:
tab |> group_by(continent) |>
  summarize(diff(quantile(log10(population), c(.25,.75))))
```


(@) Load the height data set and create a vector `x` with just the male heights:

```{r, eval=FALSE}
library(dslabs)
x <- heights$height[heights$sex=="Male"]
```

What proportion of the data is between 69 and 72 inches (taller than 69, but shorter or equal to 72)? Hint: use a logical operator and `mean`.

